# Sequence classifier #

## May 15th 2014 ##
Seek to classify sequences but have to use relevant indicators.
```{r}
setwd("~/Documents/ENS/M1/Stage/para/results/analysis/")
seqs = read.table("../22may14classify.txt", h=T)
summary(seqs)

library(ggplot2)
seqs.plots = ggplot(seqs, aes(x=length, y=GCm, label=Seq))
seqs.plots + geom_point(alpha=0.5) + labs(x="Motif Length", y="Total GC Content")
```

Let's try clustering method, without sequence column. Use of `dist` to compute distances between clusters
```{r fig.width=9, fig.height=9}
hc.complete = hclust(dist(scale(seqs[c(1:23)])), method="complete")
plot(hc.complete)
hc.cut = cutree(hc.complete, 25)
plot(hc.complete, labels=hc.cut)
rect.hclust(hc.complete,25, border=1:25)
```
We cut in 25 clusters, look like we have a good cluster #1
```{r}
seqs[hc.cut == 1,]
hc.average = hclust(dist(scale(seqs[1:23])), method="average")
plot(hc.average)
```
If we try to color the dendrogram according to cuts
```{r}
library(RColorBrewer)
source("http://addictedtor.free.fr/packages/A2R/lastVersion/R/code.R")
par(bg="#FFFFFF")
colorSet = brewer.pal(12, "Paired")
#colorSet = rainbow(20)
hc.complete$labels = new.cut
A2Rplot(hc.complete, k = 12, boxes = FALSE, col.up = "black", col.down = colorSet, knot.pos="mean", show.labels=F)

# Extend color palette
extSet = colorRampPalette(brewer.pal(9,"Set1"))(25)

# Randomize order
extSet = sample(extSet, 25, replace=F)
hc.cut = cutree(hc.complete, 20)
hc.complete$labels=hc.cut
A2Rplot(hc.complete, k = 20, boxes = FALSE, col.up = "black", col.down = extSet, knot.pos="mean", show.labels=F, lwd.down=3)

# With average cluster
extSet = sample(extSet, 20, replace=F)
hc.avg = cutree(hc.average, 25)
hc.average$labels=hc.avg
A2Rplot(hc.average, k = 25, boxes = FALSE, col.up = "black", col.down = extSet, knot.pos="mean", show.labels=F, lwd.down=3)

# Try several plots
test.list = c(10,15,20,25)
for (i in test.list) {
  A2Rplot(hc.complete, k = i, boxes = FALSE, col.up = "black", col.down = rainbow(i), knot.pos="mean", show.labels=F)
}


plot(as.dendrogram(hc.complete, hang=-20))

new.cut = cutree(hc.complete, 11)
labelColors = brewer.pal(11, "Paired")
hc.complete$labels = new.cut
# function to get color labels
colLab <- function(n) {
    if (is.leaf(n)) {
        a <- attributes(n)
        labCol <- labelColors[new.cut[which(names(new.cut) == a$label)]]
        attr(n, "nodePar") <- c(a$nodePar, lab.col = labCol)
    }
    n
}
# using dendrapply
clusDendro = dendrapply(as.dendrogram(hc.complete), colLab)
# make plot
plot(clusDendro, main = "Cool Dendrogram")
```

We could try to show a heatmap of all motifs
```{r}
seqs.mat = data.matrix(seqs)
heatmap(seqs.mat[,c(1:2, 4:23)], col = heat.colors(256), scale="column")
```

May be we should try with `vegan` package?

```{r}
library(vegan)
kclus <- kmeans(seqs[1:23], centers=11)
seqs_dist <- dist(seqs[1:23])
cmd <- cmdscale(seqs_dist)
groups <- levels(factor(kclus$cluster))
ordiplot(cmd, type = "n")
cols <- brewerpal(10, "Paired")
for(i in seq_along(groups)){
  points(cmd[factor(kclus$cluster) == groups[i], ], col = cols[i], pch = 16)
}

# add spider and hull
ordispider(cmd, factor(kclus$cluster), label = TRUE)
ordihull(cmd, factor(kclus$cluster), lty = "dotted")
```