# Ribosomal motifs #

Preliminary study of motifs in upstream regions of ribosomal proteins.
Idea: look if motifs have different distance from start codon distribution for each species, if they have, it may indicate a change in expression or function of the concerned genes?

```{r}
# Set working directory
setwd("~/Documents/ENS/M1/Stage/para/results/analysis")
```

Data loading and preparation step
```{r}
load.distribution <- function(motif.find.file, species) {
  table = read.table(motif.find.file, h=FALSE)
  colnames(table) = c("gene_id", "value", "motif")
  result = cbind(table, rep(species, length(table[,1])))
  colnames(result)[4] = "species"
  result
}
```
Load files
```{r}
bi.find = load.distribution("../ribo/bi.find.motif1", "bi")
head(bi.find)
ca.find = load.distribution("../ribo/ca.find.motif1", "ca")
sex.find = load.distribution("../ribo/sex.find.motif1", "sex")
tet.find = load.distribution("../ribo/tet.find.motif1", "tet")
data.find = data.frame(rbind(bi.find, ca.find, sex.find, tet.find))
```
How many different motifs do we have?
```{r}
unique(data.find$motif)
```
Only six different motifs are according to the matrix of the searched one.
Let's now have a look at the distributions
```{r}
library(ggplot2)
library(scales)
g = ggplot(data=data.find, aes(x=value))
g + geom_histogram(aes(fill=species), colour="black", size=1)
g + geom_density(aes(colour=species, fill=species), size=1, alpha=0.5) +
  scale_y_continuous(labels=percent)
g + geom_bar(aes(fill=species), position="dodge")
g + geom_freqpoly(aes(colour=species), size=2, alpha=0.7)
g2 = ggplot(data=data.find, aes(x=species, y=value, fill=species))
g2 + geom_boxplot(outlier.size=0) + geom_jitter(height=0) +
  scale_y_log10()
g2 + geom_violin() + geom_jitter(height=0) +
  scale_y_log10()
```

The visual comparison does not seem to show any significant difference between distributions.
We could use an ANOVA to see test this difference. But first, we have to test the homoscedasticity
```{r}
bartlett.test(value~species, data=data.find)
```

Variances are significantly different -> cannot compute an ANOVA, must use Kruskal-Wallis test. (Our samples have different sizes)

```{r}
kruskal.test(value ~ species, data=data.find)
```

Not significant -> cannot show a difference between populations

If we still compute an Anova what happens?
```{r}
model <- lm(value ~ species, data = data.find)
Anova(model)
```
Still non significant test. If we exclude the lowest sample size (*P. tetraurelia*), maybe we would see a significant difference?

```{r}
red <- subset(data.find, species != "tet")
redmodel <- lm(value ~ species, data = red)
Anova(redmodel)
```

Nothing significant for the first tested motif.

## Everything in a single function ##
To redo all analyses on other motifs let's put everything on a single function.
```{r}
# This function accept a data frame with given above columns
library(scales)
library(ggplot2)
library(car)
dist.motif <- function(data.fr) {
  unique(data.fr$motif)
  g = ggplot(data=data.fr, aes(x=value))
  print(g + geom_histogram(aes(fill=species), colour="black", size=1))
  print(g + geom_density(aes(colour=species, fill=species), size=1, alpha=0.5) +
    scale_y_continuous(labels=percent))
  print(g + geom_bar(aes(fill=species), position="dodge"))
  print(g + geom_freqpoly(aes(colour=species), size=2, alpha=0.7))
  g2 = ggplot(data=data.fr, aes(x=species, y=value, fill=species))
  print(g2 + geom_boxplot(outlier.size=0) + geom_jitter(height=0) +
    scale_y_log10())
  print(g2 + geom_violin() + geom_jitter(height=0) +
  scale_y_log10())
  # Now statistical tests
  # Homoscedasticity
  print(bartlett.test(value~species, data=data.fr))
  # Distributions comparison
  print(kruskal.test(value ~ species, data=data.fr))
  # Anova
  model <- lm(value ~ species, data = data.fr)
  Anova(model)
}
```

If we test our thing for second motif

```{r}
bi.find2 = load.distribution("../ribo/bi.find.motif2", "bi")
ca.find2 = load.distribution("../ribo/ca.find.motif2", "ca")
sex.find2 = load.distribution("../ribo/sex.find.motif2", "sex")
tet.find2 = load.distribution("../ribo/tet.find.motif2", "tet")
data.motif2 = data.frame(rbind(bi.find2, ca.find2, sex.find2, tet.find2))
dist.motif(data.motif2)
```